---
title: "Practical Aspects of Predictive Models"
author: 'Nick Hanewinckel, FSA, CERA, AVP and Actuary, HannoverRe'
output:
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: console
---
```{r setup, echo=FALSE, include=FALSE}
library(data.table)
library(ggplot2)
library(caret)
library(splines)
library(knitr)
library(kableExtra)
```

#Build a Simple Model

We will use ILEC data.

**Note to users**

The path in the code below will not work on your computer. Please download and re-map if you'd like to play along at home. The ILEC data is available:

http://cdn-files.soa.org/research/2009-15_Data_20180601.zip


```{r, readILEC}

pth <- 'H:/Mortality Research/ILEC_2009-15 Data 20180601.txt'
dat <- fread(pth,stringsAsFactors = TRUE,nrows=1000000,check.names = TRUE)
setnames(dat,'Number.of.Deaths','Deaths')
setnames(dat,'Policies.Exposed','Exposure')
dat <- dat[Exposure>0]
summary(dat)
set.seed(12345)
ind <- createDataPartition(dat$Attained.Age,p=.3,list=FALSE)
dat[,Set:='Training']
dat[ind,Set:='Testing']

```


```{r simple model}

mod1 <- glm(formula=Deaths~(bs(Attained.Age,knots = c(25,62)) + Duration + Smoker.Status)^2 + Face.Amount.Band,
            offset=log(Exposure),family = poisson,data=dat[Set=='Training'])
summary(mod1)

dat[,ExpectedDeaths:=predict(mod1,newdata=dat,type='response')]
```


## But is this safe?

```{r Safer Predict}
dat[,.(minAttAge=min(Attained.Age),maxAttAge=max(Attained.Age)),by=Set]
```

This could have been very *unsafe* if our Testing set was outside our Training set range. Without spline specifications or a saved spline object, different range. Don't leave safety to chance!

## Evaluate

```{r Evaluate}
dat[,.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set)]
dat[,.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Gender)]
dat[,.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Smoker.Status)]
dat[Attained.Age %between% c(55,75),.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Attained.Age)][order(Attained.Age)]
```

### Even Better

kableExtra gives pretty table formats!

```{r Eval Better}

kable(
  dcast(dat[,.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Gender)],Gender ~ Set,value.var=c('AtoEcount','Exposure')),digits=3)
kable(
  dcast(dat[,.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Smoker.Status)],Smoker.Status ~ Set,value.var=c('AtoEcount','Exposure')),digits=3)
kable(
  dcast(dat[Attained.Age %between% c(55,75),.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Attained.Age)][order(Attained.Age)],Attained.Age ~ Set,value.var=c('AtoEcount','Exposure')),digits=3)
```


# So...What's wrong with my model?

## Gender!

Gender was never given as a variable, but is obviously crucial. Don't get so lost in the technical details that you forget your actuarial expertise!

```{r Better Model}
mod2 <- glm(formula=Deaths~(bs(Attained.Age,knots = c(25,62)) + Duration + Smoker.Status + Gender)^2 + Face.Amount.Band,
            offset=log(Exposure),family = poisson,data=dat[Set=='Training'])
summary(mod2)

dat[,ExpectedDeaths:=predict(mod2,newdata=dat,type='response')]
dcast(dat[,.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Gender)],Gender ~ Set,value.var=c('AtoEcount','Exposure'))
dcast(dat[,.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Smoker.Status)],Smoker.Status ~ Set,value.var=c('AtoEcount','Exposure'))
dcast(dat[Attained.Age %between% c(55,75),.(Exposure=sum(Exposure),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),by=.(Set,Attained.Age)][order(Attained.Age)],Attained.Age ~ Set,value.var=c('AtoEcount','Exposure'))
```

# Graphs with Count CI
Simple example for 90% CI (5% each tail)

```{r Graph Count}
plotTable <- dat[Attained.Age %between% c(55,75),
                 .(Exposure=sum(Exposure),ExpectedDeaths=sum(ExpectedDeaths),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),
                 by=.(Set,Attained.Age)][order(Attained.Age)]

plotTable[,upperCI:=qpois(p=.95,lambda=ExpectedDeaths,lower.tail = TRUE)/ExpectedDeaths]
plotTable[,lowerCI:=qpois(p=.95,lambda=ExpectedDeaths,lower.tail = FALSE)/ExpectedDeaths]

ggplot(data=plotTable,aes(x=Attained.Age,y=AtoEcount,ymin=lowerCI,ymax=upperCI)) + 
  geom_point(aes(size=Exposure))+
  geom_errorbar()+
  facet_grid(Set~.)+
  labs(title='A/E by Attained Age - Testing vs Training',
       subtitle='90% CI')

plotTable <- dat[Attained.Age %between% c(55,75),
                 .(Exposure=sum(Exposure),ExpectedDeaths=sum(ExpectedDeaths),AtoEcount=sum(Deaths)/sum(ExpectedDeaths)),
                 by=.(Set,Gender,Attained.Age)][order(Attained.Age)]

plotTable[,upperCI:=qpois(p=.95,lambda=ExpectedDeaths,lower.tail = TRUE)/ExpectedDeaths]
plotTable[,lowerCI:=qpois(p=.95,lambda=ExpectedDeaths,lower.tail = FALSE)/ExpectedDeaths]

ggplot(data=plotTable,aes(x=Attained.Age,y=AtoEcount,ymin=lowerCI,ymax=upperCI,color=Gender)) + 
  geom_point(aes(size=Exposure))+
  geom_errorbar()+
  facet_grid(Set~.)+
  labs(title='A/E by Attained Age - Testing vs Training',
       subtitle='90% CI')
```



